Simple and Compositional Reification of Monadic Embedded Languages

### 2013-10-06 src2/arbb/paperThesis.tex
hacking

Are radical new parallel architectures market-feasible if they require
significant changes for programmers?  The jury is out.  In recent
years we have seen difficult-to-program chips suffer 
% \cite{cell} 
(e.g. Cell)
and
GPU vendors strive to enable more traditional programming features
\cite{fermi} (e.g. C++).  There is an increasing
tension between ease of programming and efficiency.

The tension  shows  across diverse chip markets.  For example,
small embedded devices are most power efficient if their processors and operating systems
omit programming features such as virtual memory and threads \cite{tinyos}.  
At the other end of the power spectrum, GPU's
graphics performance may suffer due to inclusion of hardware to ease
GPGPU programming.  In short, there is an opportunity cost to
including extra hardware for programmability.

@@

% We present our ongoing work on a
In pursuit of this vision, we offer a
% 
 new EDSL implementation, called \accarbb{}, that combines
existing systems, Accelerate \cite{ACCELERATEDAMP11} and ArBB \cite{ArBB}, to produce a unified
high-level programming environment equally suited to multicore,
vectorized CPUs, as to GPUs and other accelerators (such as Intel MIC chips \cite{larrabee}).
%
\if{0}
ArBB is based on the RapidMind \cite{RapidMind} system which was an embedded 
DSL targeting multicore CPUs as well as GPUs. ArBB does not \new{yet} include a GPU 
backend like RapidMind, targeting multicore CPUs and their vector 
units. 
\fi{}
\systemname{} is a single EDSL implementation 
with independently optimized backends \new{by different teams; 
namely, the ArBB backend for CPU/MIC, and a CUDA backend for NVIDIA GPU.}
This makes
\systemname{} an appealing platform for fair CPU/GPU comparisons, 
as well as a compelling programming model for single-source portable performance across a range of
parallel architectures, \new{present and future}.
%%To our knowledge, \systemname{} is the first EDSL implementation with
%%independently optimized backends for CPUs and GPUs.  This makes
%%\systemname{} an appealing platform for fair CPU/GPU comparisons, 
%%as well as a compelling programming
%%model for single-source portable performance across a range of
%%parallel architectures.

@@

Domain-specific languages---from Makefiles and \LaTeX{} to
Matlab---are almost too ubiquitous to notice.  Most relevant to our
purposes, domain-specific languages (DSLs) that target a narrow domain and expose communication
patterns to the compiler 
 have achieved performance-portability across a wide range of parallel
architectures.  
StreamIt\cite{streamit} is a good example.
% (for example, StreamIt\cite{streamit})
% The high-water mark in this area is perhaps  StreamIt\cite{streamit}.
%

% It has been noted \cite{not-sure} that 
% Yet the few DSLs that gain popularity lose simplicity\cite{not-sure}.  
% Yet popular DSLs rarely remain simple \cite{not-sure}.  
DSLs may start out simple and focused, but if they gain popularity
they quickly grow in complexity to rival full-blown languages.
Feature creep
can make DSL implementations complex and expensive to maintain.
Further, non-standard DSL syntax and features present a learning curve for users.  In the
last ten years an attractive solution to this dilemma has emerged: {\em
  embed} each DSL into a general-purpose host language that can provide
common functionality with familiar syntax and semantics.
% in place of the DSL.

@@

A common arrangement is for the host language program to execute at
runtime but to generate ASTs that are executed by a just-in-time
(DSL) compiler.  This use of metaprogramming (program generation)
differs from the more common usage of preprocessors and 
% Lisp/Scheme macros \cite{scheme-macros}
macros, which typically add extra
phases of computation {\em before}  compile time---increasing
the number of compile-time rather than runtime phases.

@@

One reason that EDSLs are good for productivity is that the programmer
gains the software engineering benefits of the host language
(object-orientation, higher-order-functions, modules, etc), while not
paying the cost at runtime for additional layers of abstraction or
indirection.  Indeed, the embedded languages for 
performance-oriented EDSLs are often simple, first-order
languages without pointers \cite{wavescript, ACCELERATEDAMP11}.

@@

% regiment,
As a research area, EDSLs and two-stage DSLs have been actively pursued for at least a
decade \cite{VERTIGO, wavescript} but are gaining steam
recently 
%\cite{sejits, stanford-ppl} 
\cite{copperhead, stanford-ppl} 
and are beginning to appear in
commercial products \cite{ArBB}.
%
Further, EDSL techniques have spread beyond their origin in 
 the programming languages community.
% 
For example, both Stanford's Parallel Programming Laboratory (PPL) and
the Berkeley Parlab are creating EDSLs as their flagship parallel
programming solutions for domains such as machine learning and
rendering \cite{stanford-ppl}.
  Moreover, EDSLs need not be hosted by esoteric research languages---Intel's
 ArBB embeds an array language in C++ and Berkeley's
Copperhead \cite{copperhead} generates CUDA from simple Python code.

@@

%% {\em Accelerate} (Data.Array.Accelerate) is an existing EDSL for 
%% high-level data-parallel programming in Haskell~\cite{Accelerate}. 
The Accelerate programming model consists of collective 
operations that can be performed over arrays, together with a simple 
language of scalar expressions\new{---in the current release, the Haskell type system enforces that parallelism not be nested.}
\if{0}
The Haskell type 
system is used to enforce that no nested data-parallelism can be expressed 
\new{(in the current release, this may be relaxed in the future).}
\fi{}
Accelerate's collective operations include {\em Map} (akin to parallel
for loops) {\em ZipWith} (a generalization of elementwise vector addition) and 
{\em Fold} (sum generalized)---familiar operations for programmers versed in the functional paradigm.
All of these collective operations are easily parallelizable. 
% In the case of {\em Map} and {\em ZipWith} the potential for parallelism is abundant.

@@

Using Accelerate and ArBB together, we propose a layered architecture
for \systemname{},
% Our goal is to make the benefits of ArBB available to the Haskell programmer. 
% \js{(Or is our goal to give an even better ``front-end'' to ArBB compared to C++)}
% We propose that this is best accomplished using a layered approach, as
pictured in Figure~\ref{f:architecture}. 
The host language execution (via Haskell in this case) executes the
programmer's source code, generating a vector program in the
restricted language of the Accelerate front-end.
Then either Accelerate backend---ArBB or CUDA---may be used.
The 
bottom layer of the ArBB backend consists of a direct mapping of the ArBB virtual machine
API (VMAPI) into Haskell including one-to-one bindings for each C 
function. 
To [partially] automate the creation of these bindings we used the C2HS system \cite{C2HS}. The 
\arbbvmH{} bindings are very low-level.
%and make little use of the high-level programming methods that
% Haskell programmers want. 
The idea 
is that the \arbbvmH{} bindings should be used to implement
backends for higher-level data-parallel EDSLs.

@@

\begin{enumerate}
\item Attempt to map an Accelerate reduction directly onto an ArBB
  primitive such as \cde{add\_reduce}.
\item Apply a general reduction technique based on $log(N)$ map
  operations over successively halving array sizes.  Essentially, cut
  the array in half, combine the halves, repeat. In ~\cite{reduction}
  this approach is explained in the context of CUDA. 
\end{enumerate}

@@

Black-Scholes option pricing is a finance-related benchmark that has been used in similar
DSLs targeting GPUs \cite{ACCELERATEDAMP11, NIKOLA}. Since we are re-using the 
Accelerate front-end, we can directly use the Black-Scholes benchmark that 
is shipped with that system.  Figure \ref{fig:blackscholes} shows the
complete code listing for an Accelerate Black-Scholes function which
can be executed on GPUs or any processor targeted by ArBB.

@@

These early results demonstrates the principle that even when a kernel
executes with higher throughput on a GPU, in a particular program it
is difficult to decide whether a computation is worth moving to a GPU,
incurring extra data-movement and possibly extra JIT compilation
\cite{wheres-the-data-paper}.  Specifically, we see that the
Accelerate Black-Scholes program from Figure~\ref{fig:blackscholes}
performs better on the CPU if it executes once (even on a large window
of data) whereas the GPU would yield better performance in a sustained
series of executions.
%
Because both CPU and GPU execution may be desirable---and the
selection may be dynamic---it is beneficial to have a single source code
that is portable across both.  
% Accelerate combined with \systemname{} makes this possible.

@@

OpenCL~\cite{opencl08} is a programming model very similar to CUDA but with the aspiration 
to offer both acceleration of computations on GPUs or to multicore CPUs. 
OpenCL JIT compiles the kernels for the particular hardware available and is
in that sense similar to ArBB.  
OpenCL programs are relatively low-level and require a large amount of
boilerplate to create and invoke.  In this sense they occupy a very
different niche than Accelerate.

Microsoft Accelerator~\cite{ACCELERATOR} is an embedded language with similar 
aspirations as ArBB, that is, to target a diverse range of architectures using the 
same source code. Accelerate can be used from the C\# language or the 
functional F\# language and targets GPUs or of CPUs and their
vector units.
% However, being part of the .Net framework implies limited accessibility 
%for platforms other than Microsoft Windows.

@@

%Most work on partitioning workloads between 
Many CPU and GPU comparisons, and some CPU/GPU workload partitioners \cite{merge}, rely on
redundant hand-written versions of all kernels
% \cite{cpu-gpu-partitioning} 
(though some systems like Qilin \cite{qilin} allow a single source code).
It is difficult in this kind of scenario
to make fair comparisons, controlling for the
amount of effort put into the respective implementations.
For example, comparing unoptimized serial CPU implementations vs. GPU
ones is not informative \cite{debunking-dubey}.
In \systemname{} controlling for effort
need happen only once---both CUDA and ArBB are
independently optimized by their respective teams of engineers---not for each benchmark.

@@

This is a compiler backend research challenge.
The {\em skeletons} method mentioned in
Section~\ref{sec:accelerate-arbb} is one approach to this problem, as
are the optimizations studied in the Copperhead \cite{copperhead} and
Obsidian \cite{JSLIC} projects.
%
On the other hand, systems that rely on advanced optimizations
typically suffer to some extent from performance-predictability
problems.  Thus achieving portable, predictable performance on a wide range
of architectures---even for the simplest target languages---will
be the subject of much future work.

%{
%\small
% \bibliographystyle{plainnat}
%\bibliographystyle{IEEEtran}
%\bibliography{IEEEabrv,haskell_arbb}
%}

>>>>

Are radical new parallel architectures market-feasible if they require
significant changes for programmers?  The jury is out.  In recent
years we have seen difficult-to-program chips suffer 
% \citearbb{cell} 
(e.g. Cell)
and
GPU vendors strive to enable more traditional programming features
\citearbb{fermi} (e.g. C++).  There is an increasing
tension between ease of programming and efficiency.

The tension  shows  across diverse chip markets.  For example,
small embedded devices are most power efficient if their processors and operating systems
omit programming features such as virtual memory and threads \citearbb{tinyos}.  
At the other end of the power spectrum, GPU's
graphics performance may suffer due to inclusion of hardware to ease
GPGPU programming.  In short, there is an opportunity cost to
including extra hardware for programmability.

@@

% We present our ongoing work on a
In pursuit of this vision, we offer a
% 
 new EDSL implementation, called \accarbb{}, that combines
existing systems, Accelerate \citearbb{ACCELERATEDAMP11} and ArBB \citearbb{ArBB}, to produce a unified
high-level programming environment equally suited to multicore,
vectorized CPUs, as to GPUs and other accelerators (such as Intel MIC chips \citearbb{larrabee}).
%
\if{0}
ArBB is based on the RapidMind \citearbb{RapidMind} system which was an embedded 
DSL targeting multicore CPUs as well as GPUs. ArBB does not \new{yet} include a GPU 
backend like RapidMind, targeting multicore CPUs and their vector 
units. 
\fi{}
\systemname{} is a single EDSL implementation 
with independently optimized backends \new{by different teams; 
namely, the ArBB backend for CPU/MIC, and a CUDA backend for NVIDIA GPU.}
This makes
\systemname{} an appealing platform for fair CPU/GPU comparisons, 
as well as a compelling programming model for single-source portable performance across a range of
parallel architectures, \new{present and future}.
%%To our knowledge, \systemname{} is the first EDSL implementation with
%%independently optimized backends for CPUs and GPUs.  This makes
%%\systemname{} an appealing platform for fair CPU/GPU comparisons, 
%%as well as a compelling programming
%%model for single-source portable performance across a range of
%%parallel architectures.

@@

Domain-specific languages---from Makefiles and \LaTeX{} to
Matlab---are almost too ubiquitous to notice.  Most relevant to our
purposes, domain-specific languages (DSLs) that target a narrow domain and expose communication
patterns to the compiler 
 have achieved performance-portability across a wide range of parallel
architectures.  
StreamIt\citearbb{streamit} is a good example.
% (for example, StreamIt\citearbb{streamit})
% The high-water mark in this area is perhaps  StreamIt\citearbb{streamit}.
%

% It has been noted \citearbb{not-sure} that 
% Yet the few DSLs that gain popularity lose simplicity\citearbb{not-sure}.  
% Yet popular DSLs rarely remain simple \citearbb{not-sure}.  
DSLs may start out simple and focused, but if they gain popularity
they quickly grow in complexity to rival full-blown languages.
Feature creep
can make DSL implementations complex and expensive to maintain.
Further, non-standard DSL syntax and features present a learning curve for users.  In the
last ten years an attractive solution to this dilemma has emerged: {\em
  embed} each DSL into a general-purpose host language that can provide
common functionality with familiar syntax and semantics.
% in place of the DSL.

@@

A common arrangement is for the host language program to execute at
runtime but to generate ASTs that are executed by a just-in-time
(DSL) compiler.  This use of metaprogramming (program generation)
differs from the more common usage of preprocessors and 
% Lisp/Scheme macros \citearbb{scheme-macros}
macros, which typically add extra
phases of computation {\em before}  compile time---increasing
the number of compile-time rather than runtime phases.

@@

One reason that EDSLs are good for productivity is that the programmer
gains the software engineering benefits of the host language
(object-orientation, higher-order-functions, modules, etc), while not
paying the cost at runtime for additional layers of abstraction or
indirection.  Indeed, the embedded languages for 
performance-oriented EDSLs are often simple, first-order
languages without pointers \citearbb{wavescript, ACCELERATEDAMP11}.

@@

% regiment,
As a research area, EDSLs and two-stage DSLs have been actively pursued for at least a
decade \citearbb{VERTIGO, wavescript} but are gaining steam
recently 
%\citearbb{sejits, stanford-ppl} 
\citearbb{copperhead, stanford-ppl} 
and are beginning to appear in
commercial products \citearbb{ArBB}.
%
Further, EDSL techniques have spread beyond their origin in 
 the programming languages community.
% 
For example, both Stanford's Parallel Programming Laboratory (PPL) and
the Berkeley Parlab are creating EDSLs as their flagship parallel
programming solutions for domains such as machine learning and
rendering \citearbb{stanford-ppl}.
  Moreover, EDSLs need not be hosted by esoteric research languages---Intel's
 ArBB embeds an array language in C++ and Berkeley's
Copperhead \citearbb{copperhead} generates CUDA from simple Python code.

@@

%% {\em Accelerate} (Data.Array.Accelerate) is an existing EDSL for 
%% high-level data-parallel programming in Haskell~\citearbb{Accelerate}. 
The Accelerate programming model consists of collective 
operations that can be performed over arrays, together with a simple 
language of scalar expressions\new{---in the current release, the Haskell type system enforces that parallelism not be nested.}
\if{0}
The Haskell type 
system is used to enforce that no nested data-parallelism can be expressed 
\new{(in the current release, this may be relaxed in the future).}
\fi{}
Accelerate's collective operations include {\em Map} (akin to parallel
for loops) {\em ZipWith} (a generalization of elementwise vector addition) and 
{\em Fold} (sum generalized)---familiar operations for programmers versed in the functional paradigm.
All of these collective operations are easily parallelizable. 
% In the case of {\em Map} and {\em ZipWith} the potential for parallelism is abundant.

@@

Using Accelerate and ArBB together, we propose a layered architecture
for \systemname{},
% Our goal is to make the benefits of ArBB available to the Haskell programmer. 
% \js{(Or is our goal to give an even better ``front-end'' to ArBB compared to C++)}
% We propose that this is best accomplished using a layered approach, as
pictured in Figure~\ref{f:architecture}. 
The host language execution (via Haskell in this case) executes the
programmer's source code, generating a vector program in the
restricted language of the Accelerate front-end.
Then either Accelerate backend---ArBB or CUDA---may be used.
The 
bottom layer of the ArBB backend consists of a direct mapping of the ArBB virtual machine
API (VMAPI) into Haskell including one-to-one bindings for each C 
function. 
To [partially] automate the creation of these bindings we used the C2HS system \citearbb{C2HS}. The 
\arbbvmH{} bindings are very low-level.
%and make little use of the high-level programming methods that
% Haskell programmers want. 
The idea 
is that the \arbbvmH{} bindings should be used to implement
backends for higher-level data-parallel EDSLs.

@@

\begin{enumerate}
\item Attempt to map an Accelerate reduction directly onto an ArBB
  primitive such as \cde{add\_reduce}.
\item Apply a general reduction technique based on $log(N)$ map
  operations over successively halving array sizes.  Essentially, cut
  the array in half, combine the halves, repeat. In ~\citearbb{reduction}
  this approach is explained in the context of CUDA. 
\end{enumerate}

@@

Black-Scholes option pricing is a finance-related benchmark that has been used in similar
DSLs targeting GPUs \citearbb{ACCELERATEDAMP11, NIKOLA}. Since we are re-using the 
Accelerate front-end, we can directly use the Black-Scholes benchmark that 
is shipped with that system.  Figure \ref{fig:blackscholes} shows the
complete code listing for an Accelerate Black-Scholes function which
can be executed on GPUs or any processor targeted by ArBB.

@@

These early results demonstrates the principle that even when a kernel
executes with higher throughput on a GPU, in a particular program it
is difficult to decide whether a computation is worth moving to a GPU,
incurring extra data-movement and possibly extra JIT compilation
\citearbb{wheres-the-data-paper}.  Specifically, we see that the
Accelerate Black-Scholes program from Figure~\ref{fig:blackscholes}
performs better on the CPU if it executes once (even on a large window
of data) whereas the GPU would yield better performance in a sustained
series of executions.
%
Because both CPU and GPU execution may be desirable---and the
selection may be dynamic---it is beneficial to have a single source code
that is portable across both.  
% Accelerate combined with \systemname{} makes this possible.

@@

OpenCL~\citearbb{opencl08} is a programming model very similar to CUDA but with the aspiration 
to offer both acceleration of computations on GPUs or to multicore CPUs. 
OpenCL JIT compiles the kernels for the particular hardware available and is
in that sense similar to ArBB.  
OpenCL programs are relatively low-level and require a large amount of
boilerplate to create and invoke.  In this sense they occupy a very
different niche than Accelerate.

Microsoft Accelerator~\citearbb{ACCELERATOR} is an embedded language with similar 
aspirations as ArBB, that is, to target a diverse range of architectures using the 
same source code. Accelerate can be used from the C\# language or the 
functional F\# language and targets GPUs or of CPUs and their
vector units.
% However, being part of the .Net framework implies limited accessibility 
%for platforms other than Microsoft Windows.

@@

%Most work on partitioning workloads between 
Many CPU and GPU comparisons, and some CPU/GPU workload partitioners \citearbb{merge}, rely on
redundant hand-written versions of all kernels
% \citearbb{cpu-gpu-partitioning} 
(though some systems like Qilin \citearbb{qilin} allow a single source code).
It is difficult in this kind of scenario
to make fair comparisons, controlling for the
amount of effort put into the respective implementations.
For example, comparing unoptimized serial CPU implementations vs. GPU
ones is not informative \citearbb{debunking-dubey}.
In \systemname{} controlling for effort
need happen only once---both CUDA and ArBB are
independently optimized by their respective teams of engineers---not for each benchmark.

@@

This is a compiler backend research challenge.
The {\em skeletons} method mentioned in
Section~\ref{sec:accelerate-arbb} is one approach to this problem, as
are the optimizations studied in the Copperhead \citearbb{copperhead} and
Obsidian \citearbb{JSLIC} projects.
%
On the other hand, systems that rely on advanced optimizations
typically suffer to some extent from performance-predictability
problems.  Thus achieving portable, predictable performance on a wide range
of architectures---even for the simplest target languages---will
be the subject of much future work.

%{
%\small
\bibliographystylearbb{alpha}
\bibliographyarbb{thesis}
%}

### 2013-10-06 src2/bb/body.tex
hacking

\subsection{Introduction}
\studname{} is a computer science student working in a project involving programming 
robots in a low-level imperative language. However, \studname{} has a budding 
interest in functional programming using Haskell and has read ``The Haskell 
school of expression'' \cite{HUDAK}. He gets the idea to implement 
a language for robot control embedded in Haskell. \studname{} realises that the 
capabilities of the robot hardware are very similar to those of the 
robots that Hudak evaluates graphically in a grid world. There is however 
a very important difference; the embedded language needs to be compiled into 
some form that is understood by the robot. 
%\studname{} in a computer science student with a budding interest in functional 
%programming using Haskell. \studname{} is developing  
%an embedded domain specific language (EDSL) for robot control. Inspired 
%by the book ``the Haskell school of expression'' \cite{HUDAK} but instead of 
%evaluating his robot EDSL, \studname{} wants to compile the language to a form 
%he can run on actual robot hardware. 
%with 
%different motivations, \studname{} wants compile his EDSL for actual physical robots, 
%he sets out.

@@

\noindent\colorbox{light-gray}{
\begin{minipage}[]{0.9\linewidth}
\noindent 
Hello \studname{}
\newline \newline
\noindent I'm afraid your implementation of the monadic primitives can never 
work. The constructor {\tt Return} can take any arbitrary value that has 
nothing to do with the language you're designing. These values may be 
strings, binary trees or higher-order functions from zygomorphisms to 
histomorphisms. The same problem goes for {\tt Bind}; it has to be able 
to handle arbitrary values. It cannot possibly work! This is a very 
difficult problem and such a naive solution is bound to fail. 
\newline \newline
%% \noindent For hints towards a solution to this I suggest that you follow 
%% this \underline{link} \cite{GillBlog}. Another idea is that you refer 
%% to the article ``Generic monadic constructs for embedded languages'' 
%% \cite{Generic}.
%% \newline \newline

@@

\speak{\docname{}} Others have solved this problem before but I have
never seen a solution as simple as yours and was quite surprised that
it works. For example, in this paper \direct{\docname{} shows
  \studname{} the paper \cite{Generic}}, the authors use a
continuation monad to be able to reify the monadic constructs of their
language. If they were to implement your robot language they would use
the same \texttt{Program} data type as you do. However, they would not
expose that to the user of the language. Instead they would create a
type like this:

@@

\speak{\docname{}} That's a good question. The transformation to first
order programs is dealt with by a library called Syntactic. It is
described in a separate paper \direct{\docname{} pulls out the paper
  \cite{axelsson2012generic} from a pile of papers}. It's hard to do
an apples to apples comparison between Syntactic and your
technique. Syntactic solves a much bigger problem than what you're
doing so it is naturally more complicated.

@@

\speak{\docname{}} There are also the papers
\cite{Farmer,sculthorpe2013constrained} which solves
the problem in a manner that is more similar to yours. They also have explicit
constructors {\tt Bind} and {\tt Return} but give them a slightly different
type which guarantees that all applications of {\tt Bind} are {\em normalised},
so that all {\tt Bind}s are right-associated. Unexpectedly, this
allows them to make instances of the {\tt Monad} class but at the same time
constrain the arguments of {\tt Bind}.
Their technique is more complicated than yours, but just as with the
case of Syntactic, they solve a more general problem.

@@

\speak{\docname{}} Absolutely! Compositional definitions are nice
because they imply that there is no weird semantical interaction
between the constructs as they are defined independently. But it also
means that it should be possible to factor out the constructors
\texttt{Return} and \texttt{Bind} into a data type of their own. Then,
it could be combined with other data types using techniques like Data
Types \'{a} la Carte \cite{Wouter} or CompData \cite{CompData}. This
way, a language can be designed piece by piece.You can then select the
set of pieces required for a particular task or that suits a
particular brand of robots.

@@

%% \noindent\colorbox{light-gray}{
%% \begin{minipage}[]{0.9\linewidth}
\noindent 
I note that the data type for monadic operations cannot be composed as
freely as the other operations, but instead has to be applied
separately as a final step. Your
solution is very similar to how variable binding is handled in the
Syntactic library \cite{syntactic}.
\newline \newline

@@

%A consequence of the compositionality is that we can use techniques
%such as CompData \cite{} to decompose the definition of DSLs into
%smaller reusable building blocks. Figure \ref{fig:compdata} shows a
%compositional implementation of the monadic constructs, separately
%from any other constructs. The type \verb!Cxt! is defined in the
%package CompData and ties the recursive knot.

@@

The example language used throughout the story is small and quite
limited. But the technique presented in this paper does scale up to
larger languages and more complicated language constructs.  It is
currently used in the implementation of Obsidian, an EDSL for general
purpose GPU programming \cite{Obsidian-Expressive}. Obsidian has language
constructs which are considerably more advanced than the robot
language is this paper. One example is the sequential for-loop which
is higher-order (the type parameter {\tt t} is not relevant to the
current discussion):
\begin{small}
\begin{verbatim}
SeqFor :: EWord32 -> (EWord32 -> Program t ())
       -> Program t ()
\end{verbatim}
\end{small}
There does, however, seem to be restrictions on what kind of language
constructs the reification technique presented in this paper can deal
with. For instance, the functions from the {\tt MonadPlus} class has
resisted our attempts at adding them to an EDSL in the same way as
we've added {\tt Return} and {\tt Bind}. The exact limits of the
expressiveness of our method is currently unknown to us.

@@

%Suppose you have just bought yourself a little programmable toy robot
%and you want to design a small DSL in Haskell to program it. The robot
%can turn, move forward and has a sensor to detect whether it is facing
%an obstacle or not. Inspired by the robot language in \cite{} you come
%up with a set of primitives shown in figure
%\ref{fig:interface}. In particular, you would like to use the
%do-notation to sequence programs.

>>>>

\subsection{Introduction}
\studname{} is a computer science student working in a project involving programming 
robots in a low-level imperative language. However, \studname{} has a budding 
interest in functional programming using Haskell and has read ``The Haskell 
school of expression'' \citebb{HUDAK}. He gets the idea to implement 
a language for robot control embedded in Haskell. \studname{} realises that the 
capabilities of the robot hardware are very similar to those of the 
robots that Hudak evaluates graphically in a grid world. There is however 
a very important difference; the embedded language needs to be compiled into 
some form that is understood by the robot. 
%\studname{} in a computer science student with a budding interest in functional 
%programming using Haskell. \studname{} is developing  
%an embedded domain specific language (EDSL) for robot control. Inspired 
%by the book ``the Haskell school of expression'' \citebb{HUDAK} but instead of 
%evaluating his robot EDSL, \studname{} wants to compile the language to a form 
%he can run on actual robot hardware. 
%with 
%different motivations, \studname{} wants compile his EDSL for actual physical robots, 
%he sets out.

@@

\noindent\colorbox{light-gray}{
\begin{minipage}[]{0.9\linewidth}
\noindent 
Hello \studname{}
\newline \newline
\noindent I'm afraid your implementation of the monadic primitives can never 
work. The constructor {\tt Return} can take any arbitrary value that has 
nothing to do with the language you're designing. These values may be 
strings, binary trees or higher-order functions from zygomorphisms to 
histomorphisms. The same problem goes for {\tt Bind}; it has to be able 
to handle arbitrary values. It cannot possibly work! This is a very 
difficult problem and such a naive solution is bound to fail. 
\newline \newline
%% \noindent For hints towards a solution to this I suggest that you follow 
%% this \underline{link} \citebb{GillBlog}. Another idea is that you refer 
%% to the article ``Generic monadic constructs for embedded languages'' 
%% \citebb{Generic}.
%% \newline \newline

@@

\speak{\docname{}} Others have solved this problem before but I have
never seen a solution as simple as yours and was quite surprised that
it works. For example, in this paper \direct{\docname{} shows
  \studname{} the paper \citebb{Generic}}, the authors use a
continuation monad to be able to reify the monadic constructs of their
language. If they were to implement your robot language they would use
the same \texttt{Program} data type as you do. However, they would not
expose that to the user of the language. Instead they would create a
type like this:

@@

\speak{\docname{}} That's a good question. The transformation to first
order programs is dealt with by a library called Syntactic. It is
described in a separate paper \direct{\docname{} pulls out the paper
  \citebb{axelsson2012generic} from a pile of papers}. It's hard to do
an apples to apples comparison between Syntactic and your
technique. Syntactic solves a much bigger problem than what you're
doing so it is naturally more complicated.

@@

\speak{\docname{}} There are also the papers
\citebb{Farmer,sculthorpe2013constrained} which solves
the problem in a manner that is more similar to yours. They also have explicit
constructors {\tt Bind} and {\tt Return} but give them a slightly different
type which guarantees that all applications of {\tt Bind} are {\em normalised},
so that all {\tt Bind}s are right-associated. Unexpectedly, this
allows them to make instances of the {\tt Monad} class but at the same time
constrain the arguments of {\tt Bind}.
Their technique is more complicated than yours, but just as with the
case of Syntactic, they solve a more general problem.

@@

\speak{\docname{}} Absolutely! Compositional definitions are nice
because they imply that there is no weird semantical interaction
between the constructs as they are defined independently. But it also
means that it should be possible to factor out the constructors
\texttt{Return} and \texttt{Bind} into a data type of their own. Then,
it could be combined with other data types using techniques like Data
Types \'{a} la Carte \citebb{Wouter} or CompData \citebb{CompData}. This
way, a language can be designed piece by piece.You can then select the
set of pieces required for a particular task or that suits a
particular brand of robots.

@@

%% \noindent\colorbox{light-gray}{
%% \begin{minipage}[]{0.9\linewidth}
\noindent 
I note that the data type for monadic operations cannot be composed as
freely as the other operations, but instead has to be applied
separately as a final step. Your
solution is very similar to how variable binding is handled in the
Syntactic library \citebb{syntactic}.
\newline \newline

@@

%A consequence of the compositionality is that we can use techniques
%such as CompData \citebb{} to decompose the definition of DSLs into
%smaller reusable building blocks. Figure \ref{fig:compdata} shows a
%compositional implementation of the monadic constructs, separately
%from any other constructs. The type \verb!Cxt! is defined in the
%package CompData and ties the recursive knot.

@@

The example language used throughout the story is small and quite
limited. But the technique presented in this paper does scale up to
larger languages and more complicated language constructs.  It is
currently used in the implementation of Obsidian, an EDSL for general
purpose GPU programming \citebb{Obsidian-Expressive}. Obsidian has language
constructs which are considerably more advanced than the robot
language is this paper. One example is the sequential for-loop which
is higher-order (the type parameter {\tt t} is not relevant to the
current discussion):
\begin{small}
\begin{verbatim}
SeqFor :: EWord32 -> (EWord32 -> Program t ())
       -> Program t ()
\end{verbatim}
\end{small}
There does, however, seem to be restrictions on what kind of language
constructs the reification technique presented in this paper can deal
with. For instance, the functions from the {\tt MonadPlus} class has
resisted our attempts at adding them to an EDSL in the same way as
we've added {\tt Return} and {\tt Bind}. The exact limits of the
expressiveness of our method is currently unknown to us.

@@

%Suppose you have just bought yourself a little programmable toy robot
%and you want to design a small DSL in Haskell to program it. The robot
%can turn, move forward and has a sensor to detect whether it is facing
%an obstacle or not. Inspired by the robot language in \citebb{} you come
%up with a set of primitives shown in figure
%\ref{fig:interface}. In particular, you would like to use the
%do-notation to sequence programs.
